Boosted Regression Trees for Predictive Auto-Tuning

### 2011-12-13 paper/draft1.tex

% XXX: organisation of the paper?

% XXX: see gdoc and merge this below

\subsection{Case Study}
Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
An important approach in implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Auto-tuning requires running a large number of
sub-compilations and representative computations to {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
one must basically start from scratch to develop an auto-tuned library for
stencil computations or sparse matrix operations.

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for on-demand auto-tuning.
Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:general_autotune}
\end{figure}

\subsection{Filterbank Correlation}
\label{sec:fbcorr}

Filterbank correlation is a simple spatial image filtering operation that is
an important subroutine in many image processing applications. It has a relatively
high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}. XXX cite some computer vision papers.

>>>>

\section{Predictive Auto-Tuning}

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for empirical and predictive auto-tuning.
Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:general_autotune}
\end{figure}

XXX
Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
for novel (not already auto-tuned) argument combinations.
We call this approach {\em predictive auto-tuning}.

\subsection{Learning a Regression Model}

The heart of predictive auto-tuning is a regression model that acts as surrogate
for a hand-written hardware model or empirical timing estimates.
In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
To fit this model, we
form a training set $\mathcal{X}, \mathcal{Y}$ where
each point $x^{(i)} \in \mathcal{X}$ is a tuple $(a^{(i)}, b^{(i)}, c^{(i)})$
and each target $y^{(i)} \in \mathcal{Y}$  reflects the speed of implementation $b^{(i)}$
on inputs $a^{(i)}$ on platform $c^{(i)}$.

The effectiveness of predictive auto-tuning depends on the mapping between the raw kernel timings $t(a, b, c)$ (i.e. in seconds) and the utility $y$ associated with that timing.
Anticipating that regression involves minimizing the squared error of our predictor (see Eq.~\ref{XXXbelow}) it is important to choose $y$ so that differences of a given numerical {\em magnitude} correspond to {\em improvements} of a certain utility.
In program optimization we are interested in improving the speedup over a reference implementation $b^{(\mathrm{ref})}$, so it is natural to choose
\begin{equation}
y^{(i)}
= \log\left(\frac{\mathrm{speed}(a, b, c)}{\mathrm{speed}(a, b^{(\mathrm{ref})}, c)} \right)
= \log\left(\frac{t(a, b^{(\mathrm{ref})}, c)}{t(a, b, c)} \right)
\end{equation}

One wrinkle compared with a standard regression framework is that not all kernel implementations $b$ are {\em valid} on all input configurations.

\subsection{Regression Trees}
\label{sec:brtree}

\subsection{Boosted Regression Trees}

\section{Case Study: Filterbank Correlation}

\label{sec:fbcorr}

Filterbank correlation is a simple spatial image filtering operation that is
an important subroutine in many image processing applications. It has a relatively
high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}.
%XXX cite some computer vision papers.
